Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-10-03 manuscript/main.tex
limitations and conclusion

    $$
    \begin{aligned}
    E[S(\hat{\theta}) | y^{\ast}] = \sum_l^L S(\theta_l)w_l
    \end{aligned}
    $$

@@

    The distribution for HMC are clustered very closely to zero. Both results for the bespoke sampling method are distributed far from zero which suggests strong evidence of a lack of uniformity across all parameters. This is consistent with the inspected distriubtion of rank statistics in the previous sections.

\section{Discussion}
Overall, Hamiltonian Monte Carlo applied on the reparameterised model performed best based on the calibration and efficiency evaluation criteria. Centered parameterisation using HMC performed moderately well but struggled to produce calibrated posterior estimates for $\sigma^2$ and $\phi$ as well as generate independent samples for these parameters. Gaussian mixture approximation struggled to return calibrated posterior estimates and efficient samples for both parameteristions. However, the centered parameteristion results may improve if we increased the length of the Markov chain. This is discussed later in the limitations section.

@@

Other parameterisaions of the stochastic volatility model are available as outlined in \citet{strickland2008parameterisation}, namely non centered in scale or both non centered in scale and location. The failure of non centered parameterisaion in location for the Gaussian mixture model could be due to a variety of reasons. This may be due to a particular software implementation when it comes to the Kalman Filter and Simulation Smoother (in which a specific software package was applied) that was not designed to handle the non centered parameterisation. Or perhaps this particular parameterisation doesn't suit the sampling strategy applied by KSC and another approach (e.g. HMC) might be more successful. Lots of different design and implementation details could be explored further. Pinpointing the reason why this model failed as well as exploring other configurations and MCMC algorithms in this context is left to future research.

@@

\subsection{Limitations}
- SBC is computationally expensive.

- Datasets generated dependent on the priors. If priors are diffuse then can get a wide set of possible datasets which make it harder to get calibrated estimates of the true parameter.

- Fair comparison between 999 samples and 9,999 samples.

\section{Conclusion}

>>>>

    $$
    \begin{aligned}
    E[S(\hat{\theta}) | y^{\ast}] = \sum_l^L S(\hat{\theta}_l)w_l
    \end{aligned}
    $$

@@

    The distribution for HMC are clustered very closely to zero. Both results for the bespoke sampling method are distributed far from zero which suggests strong evidence of a lack of uniformity across all parameters. This is consistent with the distribution of rank statistics in the previous sections.

\section{Discussion}
Overall, Hamiltonian Monte Carlo applied on the reparameterised model performed best based on the calibration and efficiency evaluation criteria. Centered parameterisation using HMC performed moderately well but struggled to produce calibrated posterior estimates for $\sigma^2$ and $\phi$ as well as generate independent samples for these parameters. Gaussian mixture approximation struggled to return calibrated posterior estimates and efficient samples for both parameteristions. However, the centered parameteristion results for the Gaussian mixture model may improve if we increased the length of the Markov chain. This is discussed later in the limitations section.

@@

Other parameterisaions of the stochastic volatility model are available as outlined in \citet{strickland2008parameterisation}, namely non centered in scale or both non centered in scale and location. The failure of non centered parameterisaion in location for the Gaussian mixture model could be due to a variety of reasons. This may be due to a particular software implementation when it comes to the Kalman Filter and Simulation Smoother (in which a specific software package was applied) that was not designed to handle the non centered parameterisations. Or perhaps this particular parameterisation doesn't suit the sampling strategy applied by KSC and another approach (e.g. HMC) might be more successful. Lots of different design and implementation details could be explored further. Pinpointing the reason why this model parameterisation failed as well as exploring other configurations and MCMC algorithms in this context is left to future research.

It is worth noting that there are many ways of "comparing" MCMC approaches. Other features that may be of interest is the speed in which a MCMC can converge onto the target distribution and generate samples for inference. However, "speed" is diffiuclt to define and is subject to a variety of different factors such as hardware, operating system, etc. The purpose of this research is to compare algorithms based on calibration and efficiency. Additionally, this simulation study does not provide any advice on the appropriateness of the stochastic volatility model on real data. Whether the stochastic volatility model is appropriate for modelling a particular financial time series requires further study into the data generating process, domain expertise and a suite of other diagnostic checks. Examples of this are posterior predictive checking and out of sample predictive performance. Rather, ths SBC approach gives insight into how well calibrated a sampling strategy is conditional on a known data generating process which is important in all applied contexts.

@@

\section{Limitations}
A limitation to applying SBC in all modelling contexts is that it is computationally intensive. SBC requires fitting multiple models to get reliable estimates. In some cases, fitting just one model can take a long time, let alone multiple. Producing results for this research within a reasonable timeframe required use of a computer cluster. Computing infrastructure may limit the opportunity for SBC to be performed,although it is probably not necessary to do 5000 SBC iterations in other research contexts. The choice of SBC iterations for this research was arbitrary and was used to demonstrate that noisy estimates could be reduced by increasing the number of iterations for calibrated models. A more principlied approach to picking the number of SBC iterations could be explored in other research.

Developing diagnostic tools for evaluating SBC results and models with many parameters is an area for improvement. As discussed, inspecting many histograms for highly parameterised models is unrealistic. Other summary statistics and visualisations can be applied for a high level comparison, such as the chi squared statistics applied in this research. Furthermore, it is not always straightforward to understand why any any one parameter may produce poor calibration results, for example, an arbitrary state parameter. SBC may tell us some information about the miscalibration based on the histogram shape (e.g. under or over dispersed relative to the prior distribution), but understanding why this occurs may not be straight forward. Although, diagnosing sampling problems with complicated posterior geometries speaks more generally to the difficulty of understanding complex models in the first place as opposed to a limitation of SBC.

Lastly, it is unclear what constitutes a "fair comparison" between algorithms when comparing SBC results. In particular, the number of post burnin or warmup samples differed between algorithms (999 post burn-in for HMC, 9999 for the bespoke algorithm). An argument could be made that the poor results for the bespoke algorithm may be due to the chain not converging onto the target distribution. Indeed, the authors of this approach extended their Markov chain to 750,000 posterior draws. A farier comparison may be to have roughly similar number of ESS for most of the estimated parameters, although this is difficult to get right with multiple parameters. However, given how well Hamiltoninan Monte Carlo performed with the reparamterised model, any improvement to the Gaussian mixture model by increasing the length of the Markov chain would make calibration just as good, but not better due to the inefficiency of the sampler.

\section{Conclusion}
This research evaluated and compared different MCMC algorithms for fitting stochastic volatility models. As our models and algorithms for estimating these models become more complicated, so does the need for principled ways of checking that they are returning correct posterior estimates. SBC provides a general simulation design for generative models to check whether a modelling strategy is producing the correct posterior estimates.

In the context of stochastic volatility, applying Hamiltonian Monte Carlo on a reparameterised stochastic volatility model provided the most calibrated and efficient estimates when compared to the bespoke sampling method using a Gaussian mixture model. This also outperformed SBC and ESS results when using importance sampling weights to correct for approximation error. Other parameterisations of the stochastic volatility model can be considered to see if it improves sampling performance based on these metrics.

This study only considered a handful of potential algorithms, sampling strategies and parameterisations for stochastic volatility. Future research could use the same SBC design to compare and evaluate more complicated volatility models (or any model outside this domain) and MCMC approaches. Overall, SBC can be applied as part of the development of various probabilistic and Bayesian models and is a useful tool as part of a statistical workflow.
